# Constructing a logistic regression estimator
## Dr Thomas Robinson | Day 1

In this exercise we will build a simple binary classifier from scratch, then compare it against R's built-in logistic regression. The focus is on understanding how SGD recovers the true coefficients.

## Simulate a classification problem

```{r}
set.seed(89)
n <- 500

features <- data.frame(
  X1 = runif(n, -5, 5),
  X2 = runif(n, -2, 2)
)

linear_component <- 3 + 1 * features$X1 - 2 * features$X2 + rnorm(n, 0, 0.15)
p_true <- 1 / (1 + exp(-linear_component))
y <- rbinom(n, 1, p_true)

data_all <- data.frame(y = y, features)

X <- as.matrix(cbind(Intercept = 1, data_all[, c("X1", "X2")]))

coef_true <- c(3, 1, -2)
```

## Helper functions

```{r}
sigmoid <- function(z) {
  1 / (1 + exp(-z))
}

predict_prob <- function(X, coefficients) {
  sigmoid(drop(X %*% coefficients))
}

neg_log_lik <- function(y_true, yhat_prob, eps = 1e-12) {
  yhat_prob <- pmin(pmax(yhat_prob, eps), 1 - eps)
  -mean(y_true * log(yhat_prob) + (1 - y_true) * log(1 - yhat_prob))
}
```

## A naive coefficient guess

```{r}
coef_guess <- c(0, 0.5, 1)
cat("NLL with naive guess:       ", round(neg_log_lik(y, predict_prob(X, coef_guess)), 4), "\n")
cat("NLL with true coefficients: ", round(neg_log_lik(y, predict_prob(X, coef_true)), 4), "\n")
```

## Logistic regression via stochastic gradient descent

```{r}
train_logit_sgd <- function(X, y, l_rate = 0.01, max_epochs = 200) {
  coefficients <- rep(0, ncol(X))
  history <- vector("list", max_epochs)

  for (epoch in seq_len(max_epochs)) {
    for (i in seq_len(nrow(X))) {
      row_vec <- X[i, ]
      yhat_i <- predict_prob(matrix(row_vec, nrow = 1), coefficients)
      gradient <- (yhat_i - y[i]) * row_vec
      coefficients <- coefficients - l_rate * gradient
    }

    train_prob <- predict_prob(X, coefficients)
    epoch_nll <- neg_log_lik(y, train_prob)
    history[[epoch]] <- data.frame(epoch = epoch, nll = epoch_nll)

    message(
      paste0(
        "Epoch ", epoch, "/", max_epochs,
        " | NLL = ", round(epoch_nll, 5)
      )
    )
  }

  list(coefficients = coefficients, history = do.call(rbind, history))
}
```

## Fit the model

```{r}
sgd_fit <- train_logit_sgd(X, y, l_rate = 0.02, max_epochs = 150)
round(sgd_fit$coefficients, 3)
```

## Compare coefficients and convergence

```{r}
glm_fit <- glm(y ~ X1 + X2, data = data_all, family = binomial())

rbind(
  true = coef_true,
  sgd = round(sgd_fit$coefficients, 3),
  glm = round(coef(glm_fit), 3)
)
```

```{r}
plot(
  sgd_fit$history$epoch,
  sgd_fit$history$nll,
  type = "b",
  pch = 19,
  col = "dodgerblue4",
  xlab = "Epoch",
  ylab = "Training negative log-likelihood",
  main = "SGD convergence"
)
```

The SGD model should be quite close to the built-in `glm()` estimator. That is the main point of the exercise: machine learning optimisation is often an alternative route to a familiar estimator, not a completely separate universe.

## Decision boundary

Because we only have two features, we can visualise how the classifier divides the feature space. The decision boundary is the set of points where $P(y=1) = 0.5$, which for logistic regression is a straight line.

```{r}
# Build a grid of points across the feature space
grid <- expand.grid(
  X1 = seq(-5, 5, length.out = 200),
  X2 = seq(-2, 2, length.out = 200)
)
grid_mat <- as.matrix(cbind(Intercept = 1, grid))
grid$prob <- predict_prob(grid_mat, sgd_fit$coefficients)

library(ggplot2)

ggplot() +
  geom_tile(data = grid, aes(x = X1, y = X2, fill = prob), alpha = 0.4) +
  scale_fill_gradient2(low = "firebrick", mid = "white", high = "dodgerblue3",
                       midpoint = 0.5, name = "P(y = 1)") +
  geom_point(data = data_all, aes(x = X1, y = X2, colour = factor(y)),
             size = 1.5) +
  scale_colour_manual(values = c("0" = "firebrick", "1" = "dodgerblue3"),
                      name = "Actual class") +
  geom_abline(
    intercept = -sgd_fit$coefficients[1] / sgd_fit$coefficients[3],
    slope = -sgd_fit$coefficients[2] / sgd_fit$coefficients[3],
    linewidth = 1, linetype = "dashed"
  ) +
  labs(title = "SGD decision boundary") +
  theme_minimal()
```

## Exercises

1. Change the learning rate. What happens if it is too small? What happens if it is too large?

2. Implement a stopping rule so the algorithm stops early once the loss stabilises. Hint: compare the NLL across consecutive epochs.

3. Add a third predictor `X3` that has no true effect on the outcome. Does the algorithm still learn a near-zero coefficient?

4. BONUS: implement mini-batch gradient descent so that each update uses, say, 25 observations at a time instead of a single row.
