# Constructing a logistic regression estimator
## Dr Thomas Robinson | Day 1

In this exercise we will build a simple binary classifier from scratch, then compare it against R's built-in logistic regression. The main upgrade from the earlier version is that we now use a train/test split, safer loss calculations, and an explicit stopping rule.

## Simulate a classification problem

```{r}
set.seed(89)
n <- 500

features <- data.frame(
  X1 = runif(n, -5, 5),
  X2 = runif(n, -2, 2)
)

linear_component <- 3 + 1 * features$X1 - 2 * features$X2 + rnorm(n, 0, 0.15)
p_true <- 1 / (1 + exp(-linear_component))
y <- rbinom(n, 1, p_true)

data_all <- data.frame(y = y, features)

train_idx <- sample(seq_len(nrow(data_all)), size = 0.7 * nrow(data_all))
train_data <- data_all[train_idx, ]
test_data <- data_all[-train_idx, ]

X_train <- as.matrix(cbind(Intercept = 1, train_data[, c("X1", "X2")]))
X_test <- as.matrix(cbind(Intercept = 1, test_data[, c("X1", "X2")]))
y_train <- train_data$y
y_test <- test_data$y

coef_true <- c(3, 1, -2)
```

## Helper functions

```{r}
sigmoid <- function(z) {
  1 / (1 + exp(-z))
}

predict_prob <- function(X, coefficients) {
  sigmoid(drop(X %*% coefficients))
}

neg_log_lik <- function(y_true, yhat_prob, eps = 1e-12) {
  yhat_prob <- pmin(pmax(yhat_prob, eps), 1 - eps)
  -mean(y_true * log(yhat_prob) + (1 - y_true) * log(1 - yhat_prob))
}

brier_score <- function(y_true, yhat_prob) {
  mean((y_true - yhat_prob) ^ 2)
}

accuracy <- function(y_true, yhat_prob, threshold = 0.5) {
  mean(ifelse(yhat_prob >= threshold, 1, 0) == y_true)
}

assess_classifier <- function(y_true, yhat_prob) {
  c(
    nll = neg_log_lik(y_true, yhat_prob),
    brier = brier_score(y_true, yhat_prob),
    accuracy = accuracy(y_true, yhat_prob)
  )
}
```

## A naive coefficient guess

```{r}
coef_guess <- c(0, 0.5, 1)
guess_train_prob <- predict_prob(X_train, coef_guess)
rbind(
  naive_guess = round(assess_classifier(y_train, guess_train_prob), 4),
  true_coefficients = round(assess_classifier(y_train, predict_prob(X_train, coef_true)), 4)
)
```

## Logistic regression via stochastic gradient descent

```{r}
train_logit_sgd <- function(X, y, l_rate = 0.01, max_epochs = 200, tol = 1e-05,
                            patience = 5, shuffle = TRUE) {
  coefficients <- rep(0, ncol(X))
  history <- vector("list", max_epochs)
  patience_left <- patience
  previous_nll <- Inf

  for (epoch in seq_len(max_epochs)) {
    row_order <- seq_len(nrow(X))
    if (shuffle) {
      row_order <- sample(row_order)
    }

    for (i in row_order) {
      row_vec <- X[i, ]
      yhat_i <- predict_prob(matrix(row_vec, nrow = 1), coefficients)
      gradient <- (yhat_i - y[i]) * row_vec
      coefficients <- coefficients - l_rate * gradient
    }

    train_prob <- predict_prob(X, coefficients)
    epoch_nll <- neg_log_lik(y, train_prob)
    epoch_brier <- brier_score(y, train_prob)
    history[[epoch]] <- data.frame(epoch = epoch, nll = epoch_nll, brier = epoch_brier)

    message(
      paste0(
        "Epoch ", epoch, "/", max_epochs,
        " | NLL = ", round(epoch_nll, 5),
        " | Brier = ", round(epoch_brier, 5)
      )
    )

    if (abs(previous_nll - epoch_nll) < tol) {
      patience_left <- patience_left - 1
    } else {
      patience_left <- patience
    }

    if (patience_left == 0) {
      message("Stopping early because the loss has stabilised.")
      break
    }

    previous_nll <- epoch_nll
  }

  list(coefficients = coefficients, history = do.call(rbind, history))
}
```

## Fit the model

```{r}
sgd_fit <- train_logit_sgd(X_train, y_train, l_rate = 0.02, max_epochs = 150)
round(sgd_fit$coefficients, 3)
```

## Compare coefficients and convergence

```{r}
glm_fit <- glm(y ~ X1 + X2, data = train_data, family = binomial())

rbind(
  true = coef_true,
  sgd = round(sgd_fit$coefficients, 3),
  glm = round(coef(glm_fit), 3)
)
```

```{r}
plot(
  sgd_fit$history$epoch,
  sgd_fit$history$nll,
  type = "b",
  pch = 19,
  col = "dodgerblue4",
  xlab = "Epoch",
  ylab = "Training negative log-likelihood",
  main = "SGD convergence"
)
```

## Out-of-sample performance

```{r}
sgd_test_prob <- predict_prob(X_test, sgd_fit$coefficients)
glm_test_prob <- predict(glm_fit, newdata = test_data, type = "response")
baseline_prob <- rep(mean(y_train), length(y_test))

round(
  rbind(
    baseline = assess_classifier(y_test, baseline_prob),
    sgd = assess_classifier(y_test, sgd_test_prob),
    glm = assess_classifier(y_test, glm_test_prob)
  ),
  4
)
```

The SGD model should be quite close to the built-in `glm()` estimator. That is the main point of the exercise: machine learning optimisation is often an alternative route to a familiar estimator, not a completely separate universe.

## Confusion matrix

A confusion matrix shows how predictions break down into true/false positives and negatives. This matters more than headline accuracy when classes are imbalanced or when different errors carry different costs.

```{r}
sgd_pred_class <- ifelse(sgd_test_prob >= 0.5, 1, 0)
table(Predicted = sgd_pred_class, Actual = y_test)
```

## Decision boundary

Because we only have two features, we can visualise how the classifier divides the feature space. The decision boundary is the set of points where $P(y=1) = 0.5$, which for logistic regression is a straight line.

```{r}
# Build a grid of points across the feature space
grid <- expand.grid(
  X1 = seq(-5, 5, length.out = 200),
  X2 = seq(-2, 2, length.out = 200)
)
grid_mat <- as.matrix(cbind(Intercept = 1, grid))
grid$prob <- predict_prob(grid_mat, sgd_fit$coefficients)

library(ggplot2)

ggplot() +
  geom_tile(data = grid, aes(x = X1, y = X2, fill = prob), alpha = 0.4) +
  scale_fill_gradient2(low = "firebrick", mid = "white", high = "dodgerblue3",
                       midpoint = 0.5, name = "P(y = 1)") +
  geom_point(data = test_data, aes(x = X1, y = X2, colour = factor(y)),
             size = 1.5) +
  scale_colour_manual(values = c("0" = "firebrick", "1" = "dodgerblue3"),
                      name = "Actual class") +
  geom_abline(
    intercept = -sgd_fit$coefficients[1] / sgd_fit$coefficients[3],
    slope = -sgd_fit$coefficients[2] / sgd_fit$coefficients[3],
    linewidth = 1, linetype = "dashed"
  ) +
  labs(title = "SGD decision boundary on test data") +
  theme_minimal()
```

## Exercises

1. Change the learning rate. What happens if it is too small? What happens if it is too large?

2. Turn `shuffle = FALSE`. Does training become slower or less stable?

3. Replace the stopping rule with a maximum patience over the **test** loss. Why is that a bad idea?

4. Add a third predictor `X3` that has no true effect on the outcome. Does the algorithm still learn a near-zero coefficient?

5. BONUS: implement mini-batch gradient descent so that each update uses, say, 25 observations at a time instead of a single row.
