# LASSO models and post-double selection
## Dr Thomas Robinson | Day 2

This code implements a simplified version of the post-double selection (PDS) procedure from Blackwell and Olson (2021, *Political Analysis*), applied to one of their empirical illustrations: the relationship between remittances, regime type, and political protest, originally studied by Escribà-Folch, Meseguer, and Wright (2018).

The question is whether remittances have a different effect on protest in autocracies vs. democracies. A naive single-interaction model may suffer from *omitted interaction bias* if the relationship between covariates and protest differs across regime types. PDS addresses this by using LASSO for variable selection and an unpenalised model for inference.

Note: this exercise simplifies the full PDS procedure in several ways (see exercise 4).

## Package requirements and data pre-processing

```{r}
# install.packages("haven")
# install.packages("glmnet")

library(haven)
library(glmnet)
```

First, read in the data and construct the additional variables used in the analysis:

```{r}
emw <- read_dta("../data/efmw_replication.dta")

emw$dist <- log(1 + (1 / emw$dist_coast))
emw$distwremit <- log(1 + ((emw$richremit / 1000000) * emw$dist))
```

Now restrict to the variables we need and remove rows with missing values. This is still a teaching simplification, not a recommendation for applied work — in practice you should think carefully about *why* rows are missing and what dropping them does to the target population.

```{r}
emw <- emw[, c(
  "Protest", "remit", "dict", "l1gdp", "l1pop", "l1nbr5", "l12gr",
  "l1migr", "elec3", "cowcode", "period", "distwremit",
  "caseid", "year"
)]

n_before <- nrow(emw)
emw <- na.omit(emw)
cat("Rows before na.omit:", n_before, "| after:", nrow(emw),
    "| dropped:", n_before - nrow(emw), "\n")

controls <- c("l1gdp", "l1pop", "l1nbr5", "l12gr", "l1migr", "elec3")

# Sum-to-zero contrasts avoid aliasing a single reference category with the
# intercept, which matters when the intercept is not penalised but the fixed
# effects are.  This is a common choice for penalised models.
contr.list <- list(contr.sum, contr.sum)
names(contr.list) <- c("factor(period)", "factor(cowcode)")
mod_mat <- model.matrix(
  ~ factor(period) + factor(cowcode),
  data = emw,
  contrasts.arg = contr.list
)[, -1]
```

## Define the matrices used in post-double selection

Following the paper's notation: $D$ = treatment (remittances), $V$ = moderator (autocracy indicator, called `dict` in the data), $X$ = controls, $Y$ = outcome (protest). The candidate control set $Z$ includes $V$, $X$, and the moderator--covariate interactions $VX$.

```{r}
X <- as.matrix(cbind(emw[, controls], mod_mat))
V <- emw$dict       # binary moderator: 1 = autocracy
VX <- as.matrix(V * X)
D <- emw$remit       # continuous treatment: remittances
DV <- D * V          # treatment-moderator interaction
Y <- emw$Protest     # binary outcome

RHS_matrix <- as.matrix(cbind(V = V, X, VX))
colnames(RHS_matrix) <- c("V", colnames(X), paste0("V_", colnames(X)))
```

## Stage 1. Estimate cross-validated LASSO models

PDS runs LASSO on the outcome $Y$, the treatment $D$, and the treatment--moderator interaction $DV$. Running all three — not just the outcome model — guards against *indirect regularisation bias*: a LASSO on $Y$ alone might drop covariates that weakly predict protest but strongly predict remittances, causing omitted variable bias in the interaction estimate.

```{r}
set.seed(89)

cv_lasso_selector <- function(LHS, RHS, family = c("gaussian", "binomial")) {
  family <- match.arg(family)
  type_measure <- if (family == "binomial") "deviance" else "mse"

  fit <- cv.glmnet(
    x = RHS,
    y = LHS,
    alpha = 1,
    family = family,
    type.measure = type_measure,
    nfolds = 10
  )

  selected <- which(as.vector(coef(fit, s = "lambda.1se")) != 0) - 1

  list(
    fit = fit,
    lambda = fit$lambda.1se,
    index = selected[selected > 0]
  )
}

Y_lasso <- cv_lasso_selector(LHS = Y, RHS = RHS_matrix, family = "binomial")
D_lasso <- cv_lasso_selector(LHS = D, RHS = RHS_matrix, family = "gaussian")
DV_lasso <- cv_lasso_selector(LHS = DV, RHS = RHS_matrix, family = "gaussian")

selected_columns <- sort(unique(c(Y_lasso$index, D_lasso$index, DV_lasso$index)))
selected_names <- colnames(RHS_matrix)[selected_columns]

cat("Y lambda.1se:", round(Y_lasso$lambda, 5), "\n")
cat("D lambda.1se:", round(D_lasso$lambda, 5), "\n")
cat("DV lambda.1se:", round(DV_lasso$lambda, 5), "\n")
cat("Number of selected columns:", length(selected_columns), "\n")
head(selected_names, 20)
```

The `cv.glmnet` object stores the cross-validation curve. Plotting it shows how the error changes across the regularisation path and where the two common lambda choices (`lambda.min` and `lambda.1se`) fall.

```{r}
plot(Y_lasso$fit, main = "CV curve for Y (Protest) model")
```

## Stage 2. Estimate the inference model

To follow the logic of the method more closely, we keep the base terms and then add any selected interaction-type controls from Stage 1.

```{r}
selected_interactions <- setdiff(selected_names, c("V", colnames(X)))

ds_matrix <- data.frame(
  Protest = Y,
  remit = D,
  dict = V,
  remit_dict = DV,
  as.data.frame(X),
  as.data.frame(RHS_matrix[, selected_interactions, drop = FALSE])
)

ds_matrix <- ds_matrix[, !duplicated(names(ds_matrix))]
```

Now compare the naive model against the post-double-selection model:

```{r}
ds_model <- glm(Protest ~ ., data = ds_matrix, family = binomial())

naive_model <- glm(
  Protest ~ remit * dict + l1gdp + l1pop + l1nbr5 + l12gr + l1migr + elec3 +
    factor(period) + factor(cowcode),
  data = emw,
  family = binomial()
)
```

```{r}
message("Naive model:")
coef(summary(naive_model))[c("remit", "remit:dict"), ]

message("Post-double-selection model:")
coef(summary(ds_model))[c("remit", "remit_dict"), ]
```

The key idea is that LASSO is used only as a **selection** device (Stage 1), and the final coefficients come from an unpenalised model (Stage 2). This avoids *direct regularisation bias* (LASSO shrinking the interaction of interest toward zero) and *indirect regularisation bias* (LASSO dropping covariates that matter for the treatment but not the outcome).

## Quick train/test check

The analysis above is an inference exercise, but we can still ask: does the post-double-selection model predict better than the naive model on held-out data? This reinforces the train/validation/test idea from the lecture.

```{r}
set.seed(42)
test_idx <- sample(seq_len(nrow(emw)), size = floor(0.2 * nrow(emw)))
test_rows <- emw[test_idx, ]

naive_test_prob <- predict(naive_model, newdata = test_rows, type = "response")
ds_test_prob <- predict(ds_model, newdata = ds_matrix[test_idx, ], type = "response")

brier <- function(y, p) mean((y - p)^2)

cat("Naive model  — test Brier score:", round(brier(test_rows$Protest, naive_test_prob), 4), "\n")
cat("PDS model    — test Brier score:", round(brier(test_rows$Protest, ds_test_prob), 4), "\n")
```

Note: this is a rough illustration only. Because both models were estimated on the full data, the test split is not truly held out. Exercise 3 below asks you to fix this properly.

## Exercises

1. Change `lambda.1se` to `lambda.min`. What happens to the number of selected variables?

2. Simulate a dataset with many weak predictors and only a few strong ones. Does cross-validated LASSO recover the strong signals?

3. Add a train/test split before Stage 1 and compare the predictive performance of the naive model and the post-double-selection model.

4. Read Blackwell and Olson (2021) more closely. This exercise simplifies their procedure in several ways — can you identify which? Hints: look at how the paper chooses penalty loadings (Section 3.2, equations 6--8), how it handles clustering (Section 3.4), and what variance estimator it proposes for the interaction (Section 3.3).
