# Neural Networks in R
## Dr Thomas Robinson | Day 4

This walkthrough now uses the local CCES data rather than downloading a generic benchmark dataset. That keeps the exercise self-contained and makes the comparison with earlier models more substantively meaningful for a social-science audience.

The main operational risk with Keras in R is usually the backend and Python environment, not the network code itself. So this notebook is now set up to:

* stay inside the `keras3` API
* default to a CPU backend for fewer machine-specific failures
* let students change the architecture without changing the environment logic

## System setup

```{r}
# install.packages(c("keras3", "recipes", "magrittr"))
#
# Recommended for class:
# 1. Stay on CPU unless you already know your GPU setup works.
# 2. Pick a backend once and keep it there for the whole session.
# 3. Let keras3 resolve backend dependencies from that choice.
# 4. If you want to force a backend before knitting, run for example:
#    Sys.setenv(KERAS_BACKEND = "torch")
#    # or:
#    Sys.setenv(KERAS_BACKEND = "tensorflow")
#    install_keras() is still supported, but is no longer required here.
```

## Load packages and data

```{r}
library(keras3)
library(magrittr)
library(recipes)

preferred_backend <- Sys.getenv("KERAS_BACKEND", unset = "torch")

tryCatch(
  use_backend(preferred_backend, gpu = FALSE),
  error = function(e) {
    stop(
      paste0(
        "Keras backend setup failed. Try setting one of:\n",
        "  Sys.setenv(KERAS_BACKEND = \"torch\")\n",
        "  Sys.setenv(KERAS_BACKEND = \"tensorflow\")\n",
        "then restart the R session and rerun the notebook.\n\n",
        "Original error: ", conditionMessage(e)
      ),
      call. = FALSE
    )
  }
)

clear_session()
set_random_seed(89)

message("Using Keras backend: ", config_backend())
```

```{r}
cces <- read.csv("../data/cces_formatted_oxml.csv")

for (v in seq_along(cces)) {
  if (is.character(cces[[v]])) {
    cces[[v]] <- as.factor(cces[[v]])
  }
}

cces$votetrump <- factor(ifelse(cces$vote2016 == "Donald Trump", "Trump", "Other"))

model_vars <- c(
  "birthyr", "gender", "educ", "race", "employ", "marstat",
  "pid3", "ideo5", "newsint", "faminc_new", "unionhh",
  "ownhome", "urbancity", "votetrump"
)

cces <- cces[, model_vars]
cces <- na.omit(cces)
```

## Split the data

```{r}
set.seed(89)

train_index <- sample(seq_len(nrow(cces)), size = floor(0.7 * nrow(cces)))
train_data <- cces[train_index, ]
test_data <- cces[-train_index, ]

y_train <- ifelse(train_data$votetrump == "Trump", 1, 0)
y_test <- ifelse(test_data$votetrump == "Trump", 1, 0)

baseline_pred <- rep(ifelse(mean(y_train) >= 0.5, 1, 0), length(y_test))
baseline_acc <- mean(baseline_pred == y_test)
baseline_acc
```

## Feature engineering with recipes

```{r}
rec_obj <- recipe(votetrump ~ ., data = train_data) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_numeric_predictors()) %>%
  prep()

train_processed <- bake(rec_obj, new_data = train_data)
test_processed <- bake(rec_obj, new_data = test_data)

x_train <- as.matrix(train_processed[, setdiff(names(train_processed), "votetrump")])
x_test <- as.matrix(test_processed[, setdiff(names(test_processed), "votetrump")])
```

## Construct a neural network

Now we are ready to build our neural network. The pipeline approach is similar to how you might construct a neural network in Python with `pytorch` or `tensorflow`:

```{r}
model <- keras_model_sequential() %>%
  layer_dense(units = 64, activation = "relu", input_shape = ncol(x_train)) %>%
  layer_dense(units = 32, activation = "relu") %>%
  layer_dense(units = 1, activation = "sigmoid") %>%

  # Once defined, we compile the network
  compile(
    optimizer = "adam",
    loss = "binary_crossentropy",
    metrics = c("accuracy")
  )

summary(model)
```

## Train the network

The construction of the model is a separate step from fitting. The entire network already exists at construction — training *adjusts* the weights and biases to better learn the data relationships.

```{r}
history <- fit(
  object = model,
  x = x_train,
  y = y_train,
  batch_size = 50,
  epochs = 50,
  validation_split = 0.3
)
```

## Dropout

We discussed in the lecture that we can further regularise our model by implementing dropout, a form of random noise that blocks signals passing through certain nodes. In keras, dropout is just a layer we add after a layer of conventional nodes:

```{r}
model_w_dropout <- keras_model_sequential() %>%
  layer_dense(units = 64, activation = "relu", input_shape = ncol(x_train)) %>%
  layer_dropout(rate = 0.5) %>%
  layer_dense(units = 32, activation = "relu") %>%
  layer_dropout(rate = 0.5) %>%
  layer_dense(units = 1, activation = "sigmoid") %>%

  compile(
    optimizer = "adam",
    loss = "binary_crossentropy",
    metrics = c("accuracy")
  )

history_dropout <- fit(
  object = model_w_dropout,
  x = x_train,
  y = y_train,
  batch_size = 50,
  epochs = 50,
  validation_split = 0.3
)
```

## Evaluate out-of-sample performance

```{r}
# Model without dropout
nn_pred <- ifelse(as.numeric(predict(model, x_test, verbose = 0)) >= 0.5, 1, 0)
nn_acc <- mean(nn_pred == y_test)

# Model with dropout
nn_drop_pred <- ifelse(as.numeric(predict(model_w_dropout, x_test, verbose = 0)) >= 0.5, 1, 0)
nn_drop_acc <- mean(nn_drop_pred == y_test)

data.frame(
  model = c("Baseline (majority class)", "Neural network", "Neural network + dropout"),
  accuracy = round(c(baseline_acc, nn_acc, nn_drop_acc), 4)
)
```

Compare these accuracy numbers to your random forest and BART results from Day 3. Neural networks often require more preprocessing and tuning than tree-based methods, so they should earn their place in the workflow by outperforming simpler models.

## Exercises

1. How does the neural network compare to the random forest/BART models from Day 3? Is that surprising?

2. Try to improve prediction accuracy by varying the network structure/hyperparameters (number of layers, units, dropout rate, batch size, optimizer).

3. Extend the network to a multi-class prediction problem: predict `"Trump"`, `"Clinton"`, or `"Other"`. Hint: think carefully about the final layer activation function and the loss function.
