# Neural Networks in R
## Dr Thomas Robinson | Day 4

This walkthrough now uses the local CCES data rather than downloading a generic benchmark dataset. That keeps the exercise self-contained and makes the comparison with earlier models more substantively meaningful for a social-science audience.

The main operational risk with Keras in R is usually the backend and Python environment, not the network code itself. So this notebook is now set up to:

* stay inside the `keras3` API
* default to a CPU backend for fewer machine-specific failures
* let students change the architecture without changing the environment logic

## System setup

```{r}
# install.packages(c("keras3", "recipes", "magrittr"))
#
# Recommended for class:
# 1. Stay on CPU unless you already know your GPU setup works.
# 2. Pick a backend once and keep it there for the whole session.
# 3. Let keras3 resolve backend dependencies from that choice.
# 4. If you want to force a backend before knitting, run for example:
#    Sys.setenv(KERAS_BACKEND = "torch")
#    # or:
#    Sys.setenv(KERAS_BACKEND = "tensorflow")
#    install_keras() is still supported, but is no longer required here.
```

## Load packages and data

```{r}
library(keras3)
library(magrittr)
library(recipes)

preferred_backend <- Sys.getenv("KERAS_BACKEND", unset = "torch")

tryCatch(
  use_backend(preferred_backend, gpu = FALSE),
  error = function(e) {
    stop(
      paste0(
        "Keras backend setup failed. Try setting one of:\n",
        "  Sys.setenv(KERAS_BACKEND = \"torch\")\n",
        "  Sys.setenv(KERAS_BACKEND = \"tensorflow\")\n",
        "then restart the R session and rerun the notebook.\n\n",
        "Original error: ", conditionMessage(e)
      ),
      call. = FALSE
    )
  }
)

clear_session()
set_random_seed(89)

message("Using Keras backend: ", config_backend())
```

```{r}
cces <- read.csv("../data/cces_formatted_oxml.csv")

for (v in seq_along(cces)) {
  if (is.character(cces[[v]])) {
    cces[[v]] <- as.factor(cces[[v]])
  }
}

cces$votetrump <- factor(ifelse(cces$vote2016 == "Donald Trump", "Trump", "Other"))

model_vars <- c(
  "birthyr", "gender", "educ", "race", "employ", "marstat",
  "pid3", "ideo5", "newsint", "faminc_new", "unionhh",
  "ownhome", "urbancity", "votetrump"
)

cces <- cces[, model_vars]
cces <- na.omit(cces)
```

## Split the data

```{r}
set.seed(89)

train_index <- sample(seq_len(nrow(cces)), size = floor(0.7 * nrow(cces)))
train_data <- cces[train_index, ]
test_data <- cces[-train_index, ]

y_train <- ifelse(train_data$votetrump == "Trump", 1, 0)
y_test <- ifelse(test_data$votetrump == "Trump", 1, 0)

baseline_pred <- rep(ifelse(mean(y_train) >= 0.5, 1, 0), length(y_test))
baseline_acc <- mean(baseline_pred == y_test)
baseline_acc
```

## Feature engineering with recipes

```{r}
rec_obj <- recipe(votetrump ~ ., data = train_data) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_numeric_predictors()) %>%
  prep()

train_processed <- bake(rec_obj, new_data = train_data)
test_processed <- bake(rec_obj, new_data = test_data)

x_train <- as.matrix(train_processed[, setdiff(names(train_processed), "votetrump")])
x_test <- as.matrix(test_processed[, setdiff(names(test_processed), "votetrump")])
```

## Build a neural network

```{r}
build_mlp <- function(input_dim,
                      hidden_units = c(64, 32),
                      dropout = 0.25,
                      learning_rate = 0.001) {
  model <- keras_model_sequential()

  model <- model %>%
    layer_input(shape = input_dim)

  for (units in hidden_units) {
    model <- model %>%
      layer_dense(units = units, activation = "relu")

    if (dropout > 0) {
      model <- model %>%
        layer_dropout(rate = dropout)
    }
  }

  model %>%
    layer_dense(units = 1, activation = "sigmoid") %>%
    compile(
      optimizer = optimizer_adam(learning_rate = learning_rate),
      loss = "binary_crossentropy",
      metrics = c("accuracy")
    )
}

model <- build_mlp(
  input_dim = ncol(x_train),
  hidden_units = c(64, 32),
  dropout = 0.25,
  learning_rate = 0.001
)
```

## Inspect the model

```{r}
summary(model)
```

## Train with early stopping

```{r}
history <- fit(
  object = model,
  x = x_train,
  y = y_train,
  batch_size = 256,
  epochs = 40,
  validation_split = 0.2,
  callbacks = list(
    callback_early_stopping(
      monitor = "val_loss",
      patience = 3,
      restore_best_weights = TRUE
    ),
    callback_reduce_lr_on_plateau(
      monitor = "val_loss",
      factor = 0.5,
      patience = 2
    )
  ),
  verbose = 0
)
```

```{r}
plot(history)
```

## Evaluate out-of-sample performance

```{r}
nn_test_prob <- as.numeric(predict(model, x_test, verbose = 0))
nn_test_pred <- ifelse(nn_test_prob >= 0.5, 1, 0)

accuracy <- mean(nn_test_pred == y_test)
brier <- mean((y_test - nn_test_prob) ^ 2)

data.frame(
  model = c("Baseline", "Neural network"),
  accuracy = c(baseline_acc, accuracy),
  brier = c(mean((y_test - mean(y_train)) ^ 2), brier)
)
```

This is the main applied lesson from today: neural networks often require more preprocessing and more tuning than earlier models, so they should earn their place in the workflow by improving predictive performance or by solving a representation-learning problem that simpler models cannot handle well.

## Exercises

1. Compare this neural network to your random forest and BART results from Day 3. Does the neural network clearly outperform them?

2. Change the architecture:
   * number of hidden layers
   * number of units
   * dropout rate
   * batch size

3. Try a multi-class version of the problem by predicting `"Trump"`, `"Clinton"`, or `"Other"`. What needs to change in the final layer and the loss function?

4. Calculate performance separately by `gender` or `race`. Does the model perform similarly across groups?

5. Introduce missingness into one or two predictors and compare:
   * complete-case analysis
   * simple imputation inside the train/test workflow
   * an explicit missingness indicator

Which choice looks safest substantively?
