# Ensemble Methods
## Dr Thomas Robinson | Day 5

This walkthrough combines several models from earlier in the week. The key upgrade from the earlier version is that the stacking weights are now estimated on **out-of-fold** predictions rather than in-sample fitted values, which makes the final ensemble much more credible.

## Package requirements

```{r}
# install.packages(c("glmnet", "randomForest"))

library(glmnet)
library(randomForest)
```

## Load and prepare the data

```{r}
cces <- read.csv("../data/cces_formatted_oxml.csv")

for (v in seq_along(cces)) {
  if (is.character(cces[[v]])) {
    cces[[v]] <- as.factor(cces[[v]])
  }
}

cces$votetrump <- ifelse(cces$vote2016 == "Donald Trump", 1, 0)

train_vars <- c(
  "birthyr", "gender", "educ", "race", "employ",
  "marstat", "pid3", "ideo5", "newsint"
)

set.seed(89)
train_indices <- sample(seq_len(nrow(cces)), size = floor(0.7 * nrow(cces)))

x_train <- cces[train_indices, train_vars, drop = FALSE]
x_test <- cces[-train_indices, train_vars, drop = FALSE]
y_train <- cces$votetrump[train_indices]
y_test <- cces$votetrump[-train_indices]
```

## Helper functions

```{r}
fit_logit <- function(X, y) {
  glm(votetrump ~ ., data = data.frame(votetrump = y, X), family = binomial())
}

predict_logit <- function(model, new_X) {
  as.numeric(predict(model, newdata = new_X, type = "response"))
}

fit_lasso <- function(X, y) {
  mm <- model.matrix(~ ., data = X)

  list(
    fit = cv.glmnet(
      x = mm[, -1, drop = FALSE],
      y = y,
      alpha = 1,
      family = "binomial",
      type.measure = "deviance"
    ),
    columns = colnames(mm),
    terms = attr(mm, "terms")
  )
}

predict_lasso <- function(object, new_X) {
  mm_new <- model.matrix(object$terms, data = new_X)

  missing_cols <- setdiff(object$columns, colnames(mm_new))
  if (length(missing_cols) > 0) {
    for (col in missing_cols) {
      mm_new <- cbind(mm_new, setNames(data.frame(0), col))
    }
  }

  mm_new <- mm_new[, object$columns, drop = FALSE]

  as.numeric(
    predict(
      object$fit,
      newx = mm_new[, -1, drop = FALSE],
      s = "lambda.1se",
      type = "response"
    )
  )
}

fit_rf <- function(X, y) {
  randomForest(votetrump ~ ., data = data.frame(votetrump = factor(y), X), ntree = 500)
}

predict_rf <- function(model, new_X) {
  as.numeric(predict(model, newdata = new_X, type = "prob")[, 2])
}

make_folds <- function(n, v = 5) {
  sample(rep(seq_len(v), length.out = n))
}

brier_score <- function(y_true, yhat_prob) {
  mean((y_true - yhat_prob) ^ 2)
}
```

## Estimate out-of-fold predictions

```{r}
set.seed(89)
fold_id <- make_folds(nrow(x_train), v = 5)

oof_preds <- data.frame(
  Y = y_train,
  logit = NA_real_,
  lasso = NA_real_,
  rf = NA_real_
)

for (fold in sort(unique(fold_id))) {
  valid_rows <- which(fold_id == fold)
  fit_rows <- setdiff(seq_len(nrow(x_train)), valid_rows)

  X_fit <- x_train[fit_rows, , drop = FALSE]
  X_valid <- x_train[valid_rows, , drop = FALSE]
  y_fit <- y_train[fit_rows]

  logit_fit <- fit_logit(X_fit, y_fit)
  lasso_fit <- fit_lasso(X_fit, y_fit)
  rf_fit <- fit_rf(X_fit, y_fit)

  oof_preds$logit[valid_rows] <- predict_logit(logit_fit, X_valid)
  oof_preds$lasso[valid_rows] <- predict_lasso(lasso_fit, X_valid)
  oof_preds$rf[valid_rows] <- predict_rf(rf_fit, X_valid)
}

head(oof_preds)
```

## Estimate stacking weights

We parameterise the weights so they are automatically non-negative and sum to one.

```{r}
stack_loss <- function(z, y, preds) {
  raw_w <- c(z, 0)
  weights <- exp(raw_w - max(raw_w))
  weights <- weights / sum(weights)

  mean((y - preds %*% weights) ^ 2)
}

pred_matrix <- as.matrix(oof_preds[, c("logit", "lasso", "rf")])

stack_opt <- optim(
  par = c(0, 0),
  fn = stack_loss,
  y = oof_preds$Y,
  preds = pred_matrix
)

raw_weights <- c(stack_opt$par, 0)
stack_weights <- exp(raw_weights - max(raw_weights))
stack_weights <- stack_weights / sum(stack_weights)
names(stack_weights) <- c("logit", "lasso", "rf")

stack_weights
```

## Refit constituent models on the full training data

```{r}
logit_model <- fit_logit(x_train, y_train)
lasso_model <- fit_lasso(x_train, y_train)
rf_model <- fit_rf(x_train, y_train)
```

## Test-set predictions

```{r}
test_pred_matrix <- cbind(
  logit = predict_logit(logit_model, x_test),
  lasso = predict_lasso(lasso_model, x_test),
  rf = predict_rf(rf_model, x_test)
)

stack_prob <- as.vector(test_pred_matrix %*% stack_weights)
stack_pred <- ifelse(stack_prob >= 0.5, 1, 0)

baseline_prob <- rep(mean(y_train), length(y_test))
baseline_pred <- ifelse(baseline_prob >= 0.5, 1, 0)
```

## Compare the models

```{r}
results <- data.frame(
  model = c("Baseline", "Logit", "LASSO", "Random Forest", "Stacked ensemble"),
  accuracy = c(
    mean(baseline_pred == y_test),
    mean(ifelse(test_pred_matrix[, "logit"] >= 0.5, 1, 0) == y_test),
    mean(ifelse(test_pred_matrix[, "lasso"] >= 0.5, 1, 0) == y_test),
    mean(ifelse(test_pred_matrix[, "rf"] >= 0.5, 1, 0) == y_test),
    mean(stack_pred == y_test)
  ),
  brier = c(
    brier_score(y_test, baseline_prob),
    brier_score(y_test, test_pred_matrix[, "logit"]),
    brier_score(y_test, test_pred_matrix[, "lasso"]),
    brier_score(y_test, test_pred_matrix[, "rf"]),
    brier_score(y_test, stack_prob)
  )
)

results
```

The ensemble is not guaranteed to dominate every single model on every metric in every sample, but this is the correct logic for stacking: estimate meta-weights using out-of-fold predictions, then test once on held-out data.

## Exercises

1. Add another learner to the library, such as a shallow tree or a neural network.

2. Repeat the exercise using in-sample fitted values instead of out-of-fold predictions. How much more optimistic do the ensemble results become?

3. Calculate subgroup performance for the stacked model. Does the ensemble reduce error equally across groups?

4. Change the meta-loss from mean squared error to log-loss. Do the weights change much?

5. Bin the predicted probabilities and make a simple calibration plot for each learner. Which model is best calibrated?
